---
title: "General Social Survey"
author: "Tanya Craig"
date: "Fall 2024"
output: 
  html_document:
    number_sections: true
    toc: true
---


```{r include=FALSE,echo=FALSE}
require(tigerstats)
require(tidyverse)
require(foreign)
source("GrabData.R")
```




#  Question 1

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

The data in the dataframe GSSdata is from the 2018 General Social Survey. The first blocks of R-code has selected down a subset of the data to just 16 variables. It has further removed unwanted factor levels in much of the data. Examine the code in the GrabData.R file to see what it is doing. Some of the variables are categorical and others are numerical. Be sure to do a variable analysis before tackling each question.  
First question - Is opinion on the death penalty (CAPPUN) independent of gun ownership (OWNGUN)?


## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

Question: Is gun ownership independent of opinions of capital punishment?

$H_0$: Gun ownership and opinion on capital punishment are not related.

$H_A$: Gun ownership and opinion on capital punishment are related.

OWNGUN ~ CAPPUN both are categorical, thus uses cat~cat

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results




#### Graphical Descriptive Results

```{r}
barchartGC(~OWNGUN + CAPPUN,data=GSSdata)
barchartGC(~OWNGUN + CAPPUN,data=GSSdata, type="percent")

```

This barchart shows that gun owners are more likely to favor capital punishment

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

#### Numerical Descriptive Results

```{r}
table1 <- xtabs(~OWNGUN + CAPPUN, data=GSSdata)
rowPerc(table1)
colPerc(table1)
```
This table shows the percentage of gunowners that either oppose or favor capital punishment. It looks like gun ownership is related to favoring capital punishment

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

### Inferential Results

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)


```
<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

we reject the null hypothesis because the p-value is less that 0.05. There is a relation between gun ownership and favoring capital punishment

#  Question 2

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Is there a difference in years of education (EDUC) between the those who did or did not vote in 2012 (VOTE12)?

Question: Is there a relation between those that are educated and if they did or didn't vote.
$H_O$: There is not relation between education and if you vote or not
$H_A$ : The is a relation between education and if you vote or not

Years of education is a number. Level of education is a category. Thus we use num ~ cat

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

##Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results


```{r}
boxplot(EDUC~VOTE12,data=GSSdata)
```

People who voted in 2012 seem to have more education than those who did not vote.


<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

#### Numerical Descriptive Results


```{r}
favstats(EDUC~VOTE12,data=GSSdata)
```

The table shows the average level of education of 15.72 - the mean.People who didn't vote had an average education of 13.42 years. Looks like people with higher education are more likely to vote.
<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

### Inferential Results

```{r}
t.test(EDUC~ VOTE12,data=GSSdata)
```
Based on the t-test we reject the null. 
There is a difference between the number of years of education and voting. It seems the higher the education, the more likely you are to have voted.
<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

#  Question 3

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Determine if a persons performance on the wordsum test (WORDSUM) is independent of their level of schooling (DEGREE).

Question: Does having a degree improve your scoring on the wordsum test"
$H_O$ : Having a degree does not affect your score on Wordsum
$H_A$ : Scoring on wordsum test does affect your score

Wordsum is a score, so it is a numbers. Degree is a category. We will use num ~ cat

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

##Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results
```{r}
boxplot(WORDSUM~DEGREE,data=GSSdata)
```
#### Graphical Descriptive Results
This boxplot shows that the higer your education, the better you score on wordsum.

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

#### Numerical Descriptive Results


```{r}
favstats(WORDSUM~DEGREE,data=GSSdata)
```
This graph shows that people with graduate degrees score the highest on wordsum. 

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->


```{r}
model1 <- aov(WORDSUM~DEGREE,data=GSSdata)
summary.aov(model1)
```
mosaic::t.test(WORDSUM ~ DEGREE, data = GSSdata)

Degree has more that two levels, so we use ANOVA. We reject the null hypothesis. The p=value is less than 0.05. There is a relation between having a degree and your score on wordsum


```{r}
plot(TukeyHSD(model1))
```

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->
There is no statistical evidence of a difference in highschool and less than high school. No difference between high school and jr high. There is no difference between Jr college and bachelor's; bachlor and graduate; or jr college and graduate.

